---
title       : Cognitive control of eye movements
subtitle    : A model of voluntary and involuntary saccade generation
author      : Ryan M. Hope
job         : Rensselaer Polytechnic Institute
biglogo     : rpi_logo.png
logo        : cogworks_logo.png
framework   : io2012
highlighter : highlight.js
hitheme     : tomorrow
widgets     : [mathjax]
mode        : selfcontained
knit        : slidify::knit2slides

---

```{r echo=FALSE,include=FALSE}
require(stringr)
require(pander)
require(png)
require(grid)
require(RColorBrewer)
require(ggplot2)
require(data.table)
all_summary = fread("summary_20151005131110.dat",header=T)
all_summary[,correct:=ifelse(saccade.response.target.correct,
                             paste(mode,"correct",sep="-"),
                             paste(mode,"incorrect",sep="-"))]
all_summary_clean = all_summary[average.quality==1 & !is.na(first.saccade.latency)]
r=fread("results1.txt",header=T)
r=data.table(melt(r,id.vars=1:10,measure.vars=11:28,value.name="ks.stat",variable.name="sid"))
rr=melt(r,id.vars=c("sid","ks.stat"))
r=r[,.SD[min(ks.stat)==ks.stat][1],by=sid]
r[,sid:=str_sub(sid,4)]
```

## Voluntary vs. involuntary eye movements

+ Humans make on average 1-4 saccades every second
    - thats 10s to 100s of thousands of saccades per day
    - how many are deliberately planned?
    
+ Humans feel like they can VOLUNTARILY move their eyes:
    - to any location
    - at any time
    
+ Certain circumstances absolutely produce INVOLUNTARILY saccades:
    - e.g. the sudden onset of a visual stimulus

---

## Evidence supporting automatic saccade timing

+ corrective eye movements (aka glissades)
    - short intersaccadic intervals
    - no intersaccadic interval
    - overlapping saccades

+ fixational eye movements (aka microsaccades)
    - reduce image fading from photoreceptor fatigue
    - possibly related to shifts of attention
    - follows main-sequence

+ smooth pursuit
    - lots of small regularly spaced unconscious saccades

--- .segue .dark .quote

<center><ss>How does the sense of voluntary control emerge from an oculomotor control system based on automatic saccade timing?</ss></center>

---

## CRISP DEVS (Discrete Event System Specification)

```{r fig.width=14,fig.height=6,echo=FALSE}
grid.raster(readPNG("assets/img/CRISP_DEVS.png"))
```

https://github.com/RyanHope/PyeMovements/blob/master/crisp.py

https://simpy.readthedocs.org/en/latest

---

## CRISP control mechanisms

```{r fig.width=14,fig.height=7,echo=FALSE}
grid.raster(readPNG("assets/img/crisp_plot.png"))
```

--- .segue .dark .quote

## The Experiment

--- &twocol

## Mixed-block antisaccade task

*** =left 

+ 480 trials split by 12 blocks
    - 20 anti / 20 pro per block
+ Initially `r length(all_summary[,unique(sid)])` subjects
    - Dropped 1 Sx, latencies were 1/2 that of all other Sxs
    - Dropped 4 Sxs, learned to do the task without eye movements
    - Dropped all trials with blinks

```{r echo=FALSE, fig.width=6, fig.height=2}
ggplot(all_summary_clean[!sid %in% c("e35da","06fbf","b4a6f","b608f","d17cc"),.N,by=sid]) + geom_dotplot(aes(x=N),binwidth=10) + scale_y_continuous(breaks=NULL) + ylab("") + xlab(sprintf("Trial counts the remaining %d subjects",length(all_summary_clean[!sid %in% c("e35da","06fbf","b4a6f","b608f","d17cc"),unique(sid)])))
```

*** =right

```{r fig.width=5,fig.height=6,echo=FALSE}
grid.raster(readPNG("assets/img/astask.png"))
```

---

## Eye movement parameters in the antisaccade task

```{r fig.width=14,fig.height=7,echo=FALSE}
grid.raster(readPNG("assets/img/antisaccade.png"))
```

--- .segue .dark .quote

## Experimental Results

---

## Task performance

```{r fig.width=14,fig.height=7,echo=FALSE}
ggplot(melt(all_summary_clean[!sid %in% c("e35da","06fbf","b4a6f","b608f","d17cc"),list(key.response=mean(key.response.correct),saccade.response=mean(saccade.response.target.correct)),by=c("sid","block","mode")],id.vars=1:3)[,variable2:=paste(mode,variable,sep=".")][,]) + geom_line(aes(x=block,y=value,group=variable2,color=variable2),size=1.25) + 
  facet_wrap(mode~sid,ncol=6) +
  theme(legend.position="bottom") +
  scale_x_continuous(breaks=1:12) +
  scale_colour_brewer(palette="Paired") +
  xlab("Block") +
  ylab("Response accuracy")
```

---

## First saccade latency

```{r fig.width=14,fig.height=7,echo=FALSE}
ggplot(all_summary_clean[!sid %in% c("e35da","06fbf","b4a6f","b608f","d17cc")]) + 
  geom_line(aes(x=first.saccade.latency,color=correct),stat="density",size=1.25) + 
  scale_color_manual(values=brewer.pal(name="Paired",6)[c(5,1,6,2)]) + 
  facet_wrap(~sid,ncol=6) + 
  theme(legend.position="bottom") + 
  coord_cartesian(xlim=c(0,1)) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Count")
```

---

## First saccade latency (subsetted)

```{r fig.width=14,fig.height=7,echo=FALSE}
ggplot(all_summary_clean[!sid %in% c("e35da","06fbf","b4a6f","b608f","d17cc") & (correct=="pro-correct" | correct=="anti-incorrect")]) +
  geom_line(aes(x=first.saccade.latency,color=correct),stat="density",size=1.25) + 
  scale_color_manual(values=brewer.pal(name="Paired",6)[c(1,6)]) + 
  facet_wrap(~sid,ncol=6) + 
  theme(legend.position="bottom") + 
  scale_x_continuous(breaks=seq(0,1,.25)) + 
  coord_cartesian(xlim=c(0,1)) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Count")
```

---

## First saccade amplitude

```{r fig.width=14,fig.height=7,echo=FALSE}
ggplot(all_summary_clean[!sid %in% c("e35da","06fbf","b4a6f","b608f","d17cc")]) +
  geom_point(aes(y=first.saccade.latency,x=first.saccade.amplitude.ratio,color=correct)) + 
  scale_color_manual(values=brewer.pal(name="Paired",6)[c(5,1,6,2)]) + facet_wrap(~correct,ncol=1) +
  ylab("Latency (in seconds) of first saccade post cue appearance") +
  xlab("Relative amplitude of first saccade post cue appearance")
```

--- .segue .dark .quote

## The Models

--- &twocol

## Model variants / task interaction

*** =left

Model 1: baseline
+ possible cancellation at gap
+ rate change at gap
+ possible cancellation at cue
+ rate change at cue
+ getTarget() always returns Inf

*** =right

Model 2: bottom-up
+ rate change at gap
+ rate change at cue
+ getTarget() queries environment module

Model 3: top-down
+ rate change at gap
+ attention shift at gap
+ rate change at cue
+ attention shift at cue
+ getTarget() queries attention module

---

## Model variant conceptual differences

+ <b>For the non-spatial (baseline) model, fixating is achieved via saccade cancellation.</b>

+ <b>For the spatial models, fixating is achieved via making a saccade of zero amplitude.</b>

--- &twocol

## Model: baseline | Timer rate

*** =left

```{r, echo=FALSE}
data.timer = read.delim("sim_timer.dat",header=T)
ggplot(data.timer) + 
  geom_line(aes(x=latency,group=timer_mean,color=timer_mean),stat="density",size=1.25) +
  scale_color_continuous("Mean Timer (s)") + 
  theme_bw() +
  theme(legend.position = c(0.8, 0.8)) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Density") +
  coord_cartesian(xlim=c(0,1))
```

*** =right

- The timer value seems to interact with the labile stage duration
 - When timer interval << labile duration, kurtosis <0, labile duration dominates mode
 - When timer interval ~= labile duration, kurtosis >0
 - When timer interval >> labile duration, kurtosis <0, timer interval dominates mode

---

## Model: baseline | Labile programming duration

```{r, echo=FALSE}
data.labile = read.delim("sim_labile.dat",header=T)
ggplot(data.labile) + 
  geom_line(aes(x=latency,group=labile_mean,color=labile_mean),stat="density",size=1.25) +
  scale_color_continuous("Mean Labile (s)") + 
  theme_bw() +
  theme(legend.position = c(0.8, 0.8)) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Density") +
  coord_cartesian(xlim=c(0,1))
```

---

## Model: baseline | Gap cancellation
 
```{r, echo=FALSE}
data.gap = read.delim("sim_gap.dat",header=T)
ggplot(data.gap) + 
  geom_line(aes(x=latency,group=gap_cancel_prob,color=gap_cancel_prob),stat="density",size=1.25) +
  scale_color_continuous("Gap Cancellation Probability") + 
  theme_bw() +
  theme(legend.position = c(0.8, 0.8)) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Density") +
  coord_cartesian(xlim=c(0,1))
```

---
 
## Model: baseline | Cue cancellation
 
```{r, echo=FALSE}
data.cue = read.delim("sim_cue.dat",header=T)
ggplot(data.cue) + 
  geom_line(aes(x=latency,group=cue_cancel_prob,color=cue_cancel_prob),stat="density",size=1.25) +
  scale_color_continuous("Cue Cancellation Probability") + 
  theme_bw() +
  theme(legend.position = c(0.8, 0.8)) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Density") +
  coord_cartesian(xlim=c(0,1))
```

---

## MindModeling@Home (Beta)

MindModeling@Home is volunteer computing pool dedicated to the advancement of cognitive science.

+ Based in Dayton, OH at:
    - The University of Dayton Research Institute
    - Wright State University
+ https://mindmodeling.org/
+ **Can run a year's worth of simulations in the matter of a few hours to a few days!**

Powered by:

```{r fig.width=3,fig.height=1.6,echo=FALSE}
grid.raster(readPNG("assets/img/boinc_600.png"))
```

---

## Parameter space

<table>
				<tbody><tr class="tableHead">
					<th>#</th>
					<th>Name</th>
					<th>Start</th>
					<th>Step</th>
					<th>Stop</th>
					<th>Count</th>
				</tr>
				<tr>
						<td>1.</td>
						<td>timer_states</td>
						<td>20</td>
						<td>40</td>
						<td>180</td>
						<td>5</td>
					</tr><tr>
						<td>2.</td>
						<td>timer_mean</td>
						<td>0.200</td>
						<td>0.025</td>
						<td>0.300</td> 
						<td>5					</td> 
					</tr><tr>
						<td>3.</td>
						<td>labile_mean</td>
						<td>0.120</td>
						<td>0.02</td>
						<td>0.180</td> 
						<td>4					</td> 
					</tr><tr>
						<td>4.</td>
						<td>labile_stdev</td>
						<td>8</td>
						<td>4</td>
						<td>16</td> 
						<td>3					</td> 
					</tr><tr>
						<td>5.</td>
						<td>attn_mean</td>
						<td>0.100</td>
						<td>0.025</td>
						<td>0.250</td> 
						<td>7					</td> 
					</tr><tr>
						<td>6.</td>
						<td>attn_stdev</td>
						<td>8</td>
						<td>4</td>
						<td>16</td> 
						<td>3					</td> 
					</tr><tr>
						<td>7.</td>
						<td>gap_cancel_prob</td>
						<td>0</td>
						<td>0</td>
						<td>0</td>
						<td>1</td>
					</tr><tr>
						<td>8.</td>
						<td>gap_timer_rate</td>
						<td>0.25</td>
						<td>0.25</td>
						<td>1.75</td> 
						<td>7					</td> 
					</tr><tr>
						<td>9.</td>
						<td>cue_cancel_prob</td>
						<td>0</td>
						<td>0</td>
						<td>0</td>
						<td>1</td>
					</tr><tr>
						<td>10.</td>
						<td>cue_timer_rate</td>
						<td>0.25</td>
						<td>0.25</td>
						<td>1.75</td> 
						<td>7					</td> 
					</tr><tr>
						<td>11.</td>
						<td>alpha</td>
						<td>0</td>
						<td>1</td>
						<td>1</td>
						<td>2</td>
									</tr>
			</tbody></table>

---

## Parameter space evaluation

"model run" = 500 trials

KS test = Kolmogorov–Smirnov test

+ One model run per parameter combination was dispatched out to volunteer computers
+ For each model run:
    - the resulting latency distribution (from the model) was compared to each of the 18 subjects data using a two-sample KS test
    - the two-sample KS test scores from each subject (as well as the evaluated parameter values) were returned to the MindModeling servers

+ For the following plots "best fitting" models were selected by picking the model with the lowest KS score

--- .segue .dark .quote

## MindModeling Result / Model Fits

---

## Trial Type: prosaccade | Model: bottom-up 

```{r fig.width=14,fig.height=7,echo=FALSE,warning=FALSE}
lat = fread("fits_pro-correct_bottom-up.dat",header=T)
labs = lat[,{s=ks.test(.SD[type=="data",first.saccade.latency],.SD[type=="model",first.saccade.latency]);list(ks.stat=round(s$statistic,3),p.value=round(s$p.value,3))},by=sid]
ggplot(lat) + geom_line(aes(x=first.saccade.latency,group=type,color=type,linetype=type),stat="density",size=1.25) +
  facet_wrap(~sid,ncol=6) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Count") +
  geom_text(aes(label=sprintf("ks=%.3f",ks.stat),x=.75,y=25),data=labs,size=4) +
  geom_text(aes(label=sprintf("p=%.3f",p.value),x=.75,y=21),data=labs,size=4) +
  theme(legend.position="bottom") +
  scale_linetype_manual(values=c("solid","twodash")) +
  coord_cartesian(xlim=c(0,1))
```

---

## Trial Type: prosaccade | Model: top-down

```{r fig.width=14,fig.height=7,echo=FALSE,warning=FALSE}
lat = fread("fits_pro-correct_top-down.dat",header=T)
labs = lat[,{s=ks.test(.SD[type=="data",first.saccade.latency],.SD[type=="model",first.saccade.latency]);list(ks.stat=round(s$statistic,3),p.value=round(s$p.value,3))},by=sid]
ggplot(lat) + geom_line(aes(x=first.saccade.latency,group=type,color=type,linetype=type),stat="density",size=1.25) +
  facet_wrap(~sid,ncol=6) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Count") +
  geom_text(aes(label=sprintf("ks=%.3f",ks.stat),x=.75,y=25),data=labs,size=4) +
  geom_text(aes(label=sprintf("p=%.3f",p.value),x=.75,y=21),data=labs,size=4) +
  theme(legend.position="bottom") +
  scale_linetype_manual(values=c("solid","twodash")) +
  coord_cartesian(xlim=c(0,1))
```

---

## Trial Type: antisaccade | Model: bottom-up

```{r fig.width=14,fig.height=7,echo=FALSE,warning=FALSE}
lat = fread("fits_anti-correct_bottom-up.dat",header=T)
labs = lat[,{s=ks.test(.SD[type=="data",first.saccade.latency],.SD[type=="model",first.saccade.latency]);list(ks.stat=round(s$statistic,3),p.value=round(s$p.value,3))},by=sid]
ggplot(lat) + geom_line(aes(x=first.saccade.latency,group=type,color=type,linetype=type),stat="density",size=1.25) +
  facet_wrap(~sid,ncol=6) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Count") +
  geom_text(aes(label=sprintf("ks=%.3f",ks.stat),x=.75,y=23),data=labs,size=4) +
  geom_text(aes(label=sprintf("p=%.3f",p.value),x=.75,y=19),data=labs,size=4) +
  theme(legend.position="bottom") +
  scale_linetype_manual(values=c("solid","twodash")) +
  coord_cartesian(xlim=c(0,1))
```

---

## Trial Type: antisaccade | Model: top-down

```{r fig.width=14,fig.height=7,echo=FALSE,warning=FALSE}
lat = fread("fits_anti-correct_top-down.dat",header=T)
labs = lat[,{s=ks.test(.SD[type=="data",first.saccade.latency],.SD[type=="model",first.saccade.latency]);list(ks.stat=round(s$statistic,3),p.value=round(s$p.value,3))},by=sid]
ggplot(lat) + geom_line(aes(x=first.saccade.latency,group=type,color=type,linetype=type),stat="density",size=1.25) +
  facet_wrap(~sid,ncol=6) +
  xlab("Latency (in seconds) of first saccade post cue appearance") +
  ylab("Count") +
  geom_text(aes(label=sprintf("ks=%.3f",ks.stat),x=.75,y=23),data=labs,size=4) +
  geom_text(aes(label=sprintf("p=%.3f",p.value),x=.75,y=19),data=labs,size=4) +
  theme(legend.position="bottom") +
  scale_linetype_manual(values=c("solid","twodash")) +
  coord_cartesian(xlim=c(0,1))
```

--- .segue .dark .quote

## Summary

---

## Preliminary results and conclusions

Results
+ The baseline model fits nothing
+ The bottom-up model fits:
    - pro-correct trials very well
    - anti-incorrect trials very well
    - anti-correct trials very well
+ The top-down model fits:
    - anti-correct trials

Conclusions
+ Modeling results so far support the idea that voluntary control can emerge from an oculomotor control system based on automatic saccade timing
+ Saccade cancelation is not needed to "keep the eyes from moving" (i.e. maintain fixation)

---

## Limitations and continuing work

* A large space of parameters generate latency distributions that are statistically indistinguishable from human data
    - Need to understand what this means, if any parameters are not needed
    - Need to determine if any values can be fixed for all subjects
* Need to get model performing task as mixed block instead of single blocks
    - Need to see what if any parameters work for both anti and pro trials
* Need to improve model evaluation on volunteers
    - Need to perform more than one model run and get estimate of KS stat variance

* Haven't attempted to model saccade amplitude yet
* Haven't attempted to model the latency of saccades beyond the first saccade


